A simplified dynamic bioenergetic model for coral-algal symbiosis and coral bleaching

### 2016-10-18 JTB_ms/JTB_ms.Rmd
update sensitivity analyses

![Sensitivity analysis. Plots show the fractional change in steady state values in response to fractional changes in default parameter values (see Table 2 for default values). Parameters are grouped by which processes they are involved in. Sensitivity in response to changes in each parameter was evaluated in each of four environments (indicated by line type): 1) LLLN=low light (2 mol ph m^-2^ d^-1^) and low DIN (1e-7 mol L^-1^), 2) LLHN=low light (2 mol ph m^-2^ d^-1^) and high DIN (4e-6 mol L^-1^), 3) HLLN=high light (30 mol ph m^-2^ d^-1^) and low DIN (1e-7 mol L^-1^), and 4) HLHN=high light (30 mol ph m^-2^ d^-1^) and high DIN (4e-6 mol L^-1^).](../img/Fig3.png)

>>>>

![Sensitivity analysis. Plots show the fractional change in steady state values of growth (solid lines) and S:H biomass (dashed lines) in response to fractional changes in default parameter values (see Table 2 for default values). Parameters are grouped by which processes they are involved in. This sensitivity analysis was conducted at conditions typical for coral reef environments: low DIN (1e-7 M) and intermediate light (15 mol m^-2^ d^-2^). Sensitivity analyses conducted other environmental conditions are presented in the Supplementary Information.](../img/Fig3.png)

### 2016-10-18 JTB_ms/JTB_ms.tex
update sensitivity analyses

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A simplified dynamic bioenergetic model for coral bleaching},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{A simplified dynamic bioenergetic model for coral bleaching}
    \author[University of Hawaii]{Ross Cunning\corref{c1}}
   \ead{ross.cunning@gmail.com} 
   \cortext[c1]{Corresponding Author}
    \author[University of California]{Erik B. Muller}

@@

To analyze qualitative behavior, we ran the model to steady state across
gradients of external irradiance and nutrients (Fig. 2), which revealed
patterns consistent with observed phenomena in corals. Predicted growth
rates are low at low light and DIN, and begin increasing as both of
these factors increase (Fig. 2A). Low light is limiting to
photosynthesis (Fig. 2D), which leads to carbon-limitation of host
growth (Fig. 2B) and an associated increase in the symbiont to host
biomass ratio (Fig. 2C). In agreement with this trend are many
observations of negative correlation between irradiance and symbiont
density (Stimson 1997; Brown et al. 1999; Fitt et al. 2000; Titlyanov et
al. 2001). As higher light alleviates light-limitation of
photosynthesis, growth becomes less carbon-limited. Similarly, elevating
DIN up to \textasciitilde{}2 µM alleviates nitrogen-limitation (Fig.
2B). Increased growth at higher DIN is predicted by the DEB model of
Muller et al. (2009), and has also been observed experimentally
(Muller-Parker et al. 1994; Tanaka et al. 2007; Tanaka et al. 2013).
However, DIN elevation beyond a certain point (e.g., 2-3 µM in these
simulations) has little effect on growth as carbon becomes limiting
(Fig. 2B). Very high nutrient levels may even reduce growth (Shantz,
Lemoine, and Burkepile 2015), although these impacts are not likely to
occur within the range of concentrations considered here (\textless{}4
µM) (Ferrier-Pagès et al. 2000).

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-15 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textgreater{}\textasciitilde{}3 µM DIN (Fig. 2A), are comparable to the
rate of 0.07 d\textsuperscript{-1} measured by Tanaka et al. (2007)
under similar N-enriched conditions. Growth under conditions more
typical of reef environments (\textless{}0.5 µM DIN) are within the
range of \textasciitilde{}0.01-0.05 d\textsuperscript{-1}; however,
these values are still higher than most specific growth rates reported
in the literature, which are typically based on skeletal mass, and fall
near or below 0.01 d\textsuperscript{-1} (Osinga et al. 2011; Osinga et
al. 2012) (though values as high as 0.025 d\textsuperscript{-1} have
been reported (Schutter et al. 2010)). These generally low skeletal
growth rates relative to predicted biomass growth may be explained by
differential substrate-limitation (i.e., CO\textsubscript{2} may limit
calcification, which is not present in the model); indeed, tissue and
skeletal growth may not always be well correlated (Anthony 2002).
Furthermore, bioerosion may reduce net skeletal growth, especially in
the field (Bak 1976). Nevertheless, when rates of increase in coral
tissue biomass have been directly measured over short periods, values
are consistent with model predictions (e.g., 0.07 d\textsuperscript{-1}
in Tanaka et al. (2007)), and similar rates (0.04 d\textsuperscript{-1})
have been measured in \emph{Aiptasia diaphana}, a non-calcifying
symbiotic anemone (Armoza-Zvuloni et al. 2014).

As irradiance continues to increase above \textasciitilde{}15-20 mol
photons m\textsuperscript{-2} d\textsuperscript{-1}, growth rates
decline until positive growth ceases above \textasciitilde{}30 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1}. The mechanism
underlying this decline is the increase in light energy beyond the
capacity of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing the symbiont biomass loss rate (representing photodamage
and/or symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011).
Together, these impacts reduce the symbiont to host biomass ratio (Fig.
2C), as occurs during coral bleaching. This reduction in symbionts
consequently reduces the flux of fixed carbon to the host, resulting in
increasing carbon-limitation (Fig. 2B) and eventual cessation of growth
(Fig. 2A). Importantly, these consequences arise in the present model
only when light energy exceeds the capacity of photochemistry and NPQ,
which in turn depends on CO\textsubscript{2} availability. While
previous models framed photophysiological stress as a fixed response to
absolute irradiance levels (Eynaud, Nisbet, and Muller 2011), the
current implementation considers the dynamic balance of multiple energy
sinks in the causation of stress, which is more consistent with current
understanding of symbiosis dysfunction (Wooldridge 2013), and
establishes an important role of the host in providing
CO\textsubscript{2} for symbiont photosynthesis (Tansik, Fitt, and
Hopkinson 2015; Hopkinson, Tansik, and Fitt 2015).

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig3.png}
\caption{Sensitivity analysis. Plots show the fractional change in
steady state values in response to fractional changes in default
parameter values (see Table 2 for default values). Parameters are
grouped by which processes they are involved in. Sensitivity in response
to changes in each parameter was evaluated in each of four environments
(indicated by line type): 1) LLLN=low light (2 mol ph
m\textsuperscript{-2} d\textsuperscript{-1}) and low DIN (1e-7 mol
L\textsuperscript{-1}), 2) LLHN=low light (2 mol ph
m\textsuperscript{-2} d\textsuperscript{-1}) and high DIN (4e-6 mol
L\textsuperscript{-1}), 3) HLLN=high light (30 mol ph
m\textsuperscript{-2} d\textsuperscript{-1}) and low DIN (1e-7 mol
L\textsuperscript{-1}), and 4) HLHN=high light (30 mol ph
m\textsuperscript{-2} d\textsuperscript{-1}) and high DIN (4e-6 mol
L\textsuperscript{-1}).}
\end{figure}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A simplified dynamic bioenergetic model for coral-algal symbiosis and coral bleaching},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{A simplified dynamic bioenergetic model for coral-algal symbiosis and
coral bleaching}
    \author[University of Hawaii]{Ross Cunning\corref{c1}}
   \ead{ross.cunning@gmail.com} 
   \cortext[c1]{Corresponding Author}
    \author[University of California]{Erik B. Muller}

@@

To analyze qualitative behavior, we ran the model to steady state across
gradients of external irradiance and nutrients (Fig. 2), which revealed
patterns consistent with observed phenomena in corals. Predicted growth
rates are low at low light and DIN (\textasciitilde{}0.01
d\textsuperscript{-1}), and begin increasing as both of these factors
increase (Fig. 2A). Low light is limiting to photosynthesis (Fig. 2D),
which leads to carbon-limitation of host growth (Fig. 2B) and an
associated increase in the symbiont to host biomass ratio (Fig. 2C). In
agreement with this trend are many observations of negative correlation
between irradiance and symbiont density (Stimson 1997; Brown et al.
1999; Fitt et al. 2000; Titlyanov et al. 2001). As higher light
alleviates light-limitation of photosynthesis, growth becomes less
carbon-limited. Similarly, elevating DIN up to \textasciitilde{}2 µM
alleviates nitrogen-limitation (Fig. 2B). Increased growth at higher DIN
is predicted by the DEB model of Muller et al. (2009), and has also been
observed experimentally (Muller-Parker et al. 1994; Tanaka et al. 2007;
Tanaka et al. 2013). However, DIN elevation beyond a certain point
(e.g., \textasciitilde{}3-4 µM in these simulations) has little effect
on growth as carbon becomes limiting (Fig. 2B). Very high nutrient
levels may even reduce growth (Shantz, Lemoine, and Burkepile 2015),
although these impacts are not likely to occur within the range of
concentrations considered here (\textless{}4 µM) (Ferrier-Pagès et al.
2000).

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-20 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 2A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) under similar
N-enriched conditions. Growth under conditions more typical of reef
environments (\textless{}0.5 µM DIN) are within the range of
\textasciitilde{}0.01-0.03 d\textsuperscript{-1}; these values are
slightly higher than most specific growth rates reported in the
literature, which are typically based on skeletal mass, and fall near or
below 0.01 d\textsuperscript{-1} (Osinga et al. 2011; Osinga et al.
2012) (though values as high as 0.025 d\textsuperscript{-1} have been
reported (Schutter et al. 2010)). These generally low skeletal growth
rates relative to predicted biomass growth may be explained by
differential substrate-limitation (i.e., CO\textsubscript{2} may limit
calcification, which is not present in the model); indeed, tissue and
skeletal growth may not always be well correlated (Anthony 2002).
Furthermore, bioerosion may reduce net skeletal growth, especially in
the field (Bak 1976). Nevertheless, when rates of increase in coral
tissue biomass have been directly measured over short periods, values
are consistent with model predictions (e.g., 0.07 d\textsuperscript{-1}
in Tanaka et al. (2007)), and similar rates (0.04 d\textsuperscript{-1})
have been measured in \emph{Aiptasia diaphana}, a non-calcifying
symbiotic anemone (Armoza-Zvuloni et al. 2014).

As irradiance continues to increase above \textasciitilde{}20 mol
photons m\textsuperscript{-2} d\textsuperscript{-1}, growth rates
decline until positive growth ceases above \textasciitilde{}35 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1}. The mechanism
underlying this decline is the increase in light energy beyond the
capacity of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing symbiont biomass loss (representing photodamage and/or
symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011). Together,
these impacts reduce the symbiont to host biomass ratio (Fig. 2C), as
occurs during coral bleaching. This reduction in symbionts consequently
reduces the flux of fixed carbon to the host, resulting in increasing
carbon-limitation (Fig. 2B) and eventual cessation of growth (Fig. 2A).
Importantly, these consequences arise in the present model only when
light energy exceeds the capacity of photochemistry and NPQ, which in
turn depends on CO\textsubscript{2} availability. While previous models
framed photophysiological stress as a fixed response to absolute
irradiance levels (Eynaud, Nisbet, and Muller 2011), the current
implementation considers the dynamic balance of multiple energy sinks in
the causation of stress, which is more consistent with current
understanding of symbiosis dysfunction (Wooldridge 2013), and
establishes an important role of the host in providing
CO\textsubscript{2} for symbiont photosynthesis (Tansik, Fitt, and
Hopkinson 2015; Hopkinson, Tansik, and Fitt 2015).

@@

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig3.png}
\caption{Sensitivity analysis. Plots show the fractional change in
steady state values of growth (solid lines) and S:H biomass (dashed
lines) in response to fractional changes in default parameter values
(see Table 2 for default values). Parameters are grouped by which
processes they are involved in. This sensitivity analysis was conducted
at conditions typical for coral reef environments: low DIN (1e-7 M) and
intermediate light (15 mol m\textsuperscript{-2} d\textsuperscript{-2}).
Sensitivity analyses conducted other environmental conditions are
presented in the Supplementary Information.}
\end{figure}
